The Noisy Work of Uncertainty Visualisation

### 2024-01-26 paper.qmd
formatting edits so knit works

## 2.2 Taxonomy Definitions
# 2.2.1 Taxonomies of Uncertainty
While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot. When authors try to make a considerable effort to define what we mean when we talk about uncertainty, they often express it in the form of a taxonomy.

@@

*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline. Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Refsgaard2007]. @fig-loc shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps. The relevance of every stage of the uncertainty taxonomy is backed up by both theoretical and practical work in statistics and adjacent fields. @Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore. @Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise. @Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input. High level stages such as model selection and estimation have obvious introductions of uncertainty through the choice of models and nature of estimation [^I'm not sure if I need a citation here because this feels obvious]. Location is also acknowledged as an important element of consideration in the taxonomies of several visualsiation papers [^ I will probably need a citation for which ones].

[Illustration of the connection between the data generating process and the location of the uncertainty analysis. Several examples are collected from @utypo and @Munzner2009](location.png){#fig-loc}

>>>>

## 2.2 Taxonomy Definitions
### 2.2.1 Taxonomies of Uncertainty
While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot. When authors try to make a considerable effort to define what we mean when we talk about uncertainty, they often express it in the form of a taxonomy.

@@

*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline. Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Refsgaard2007]. @fig-location shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps. The relevance of every stage of the uncertainty taxonomy is backed up by both theoretical and practical work in statistics and adjacent fields. @Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore. @Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise. @Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input. High level stages such as model selection and estimation have obvious introductions of uncertainty through the choice of models and nature of estimation [^I'm not sure if I need a citation here because this feels obvious]. Location is also acknowledged as an important element of consideration in the taxonomies of several visualsiation papers [^ I will probably need a citation for which ones].

[Illustration of the connection between the data generating process and the location of the uncertainty analysis. Several examples are collected from utypo and Munzner2009. ](location.png){#fig-location}
